---
title: "Preprocessing And Visualization"
author: "Jannis Vollmer"
date: "2024-04-05"
output: html_document
---

```{r}
  require(FSelector)
```

# Unsupervised Data Preprocessing

## Load data and get an overview
Our dataset contains primary numeric features, except the first column (country name)

```{r}
  data <- read.csv("../data sets/food.csv", row.names = 1)
  head(data)
```

## Preprocess data using z-transformation 
```{r}
  data_scale <- scale(data)
  head(data_scale)
```

## Perform a principal component analysis
```{r}
  pca <- princomp(data_scale)
```

## Plot the reduced dataset
```{r}
  data_red <- pca$scores[,1:2]
  plot(data_red[,"Comp.1"], data_red[,"Comp.2"])
```

# Supervised Data Preprocessing
## Load data and get an overview

```{r}
  data_diabetes <- read.csv("../data sets/diabetes.csv")
  data_diabetes_num <- data_diabetes[,-9]
  iqrs <- apply(data_diabetes_num, 2, function(x){IQR(x)})
  q1 <- apply(data_diabetes_num, 2, function(x){quantile(x,0.25)})
  q3 <- apply(data_diabetes_num, 2, function(x){quantile(x,0.75)})
  # if value < q1 - 3 * IQR or value > q3 + 3 * IQR
```

## Supervised Feature Selection
```{r}
  igs <- information.gain(class ~ ., data_diabetes)
  head(igs)
```

## Boxplot Visualization of one highest and one lowest
```{r}
  boxplot(plas ~ class, data_diabetes)
```